\documentclass{article}

% NeurIPS-style single-column preprint, self-contained so it compiles with pdflatex.
% To submit, drop the body into the official neurips_2024.tex (\usepackage{neurips_2024}).
\usepackage[letterpaper,margin=1.25in]{geometry}
\usepackage{times}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{microtype}
\usepackage[colorlinks=true,allcolors=blue]{hyperref}
\usepackage{xcolor}
\usepackage{titlesec}
\titleformat{\section}{\normalfont\large\bfseries}{\thesection}{0.6em}{}
\titleformat{\subsection}{\normalfont\normalsize\bfseries}{\thesubsection}{0.6em}{}
\setlength{\parindent}{0pt}\setlength{\parskip}{0.5em}
\newcommand{\KL}{\mathrm{KL}}
\newcommand{\note}[1]{\textsuperscript{\textbf{#1}}}

\title{\textbf{An Information-Bottleneck Adapter for Audio Grounding in Frozen\\
Audio--Visual LLMs: Preference-Optimization Collapse and Its Mitigation}}
\author{Anonymous}
\date{}

\begin{document}
\maketitle

\begin{abstract}
We study whether a frozen audio--visual large language model (AV-LLM) can be adapted to ground its
answers in audio rather than vision by training only a small per-modality variational information
bottleneck (VIB) on its modality-adapter outputs, using a modality-conditional (audio-swap)
preference objective. Applying direct preference optimization (DPO) to this setup causes the model to
collapse: on a held-out perception benchmark its accuracy falls from $0.953$ to $0.013$ and it
converges to a near-constant answer. We attribute this to likelihood displacement and the absence of
any constraint on the model's output distribution. We then add two anchors --- a term that holds the
chosen answer's likelihood at or above the frozen base, and an explicit KL penalty to the base on a set
of general inputs --- and an ablation shows that both are required: the first alone still loses
perception accuracy and the second alone loses hallucination-resistance. With both, on Qwen3-Omni and under a
leakage-controlled protocol (checkpoints selected on a held-out validation split, reported on a disjoint
test split), the audio--visual hallucination benchmark AVHBench improves from $0.647$ to $0.694\pm0.023$
across three seeds, while the two CMM capability axes stay within tolerance of the base. We report a sweep over the
information-bottleneck rate weight $\beta_{kl}$, a per-token analysis of the trained bottleneck, and
partial results on Qwen2.5-Omni. Results on VideoLLaMA2 are pending a numerical fix and are marked
\note{**} throughout. All gains are modest and measured on $300$-example benchmark subsets; limitations
are discussed in detail.
\end{abstract}

\section{Introduction}
Audio--visual large language models couple pretrained audio and vision encoders to a language model
through lightweight per-modality adapters. They exhibit a modality imbalance: they tend to answer from
the visual stream and under-use audio, including hallucinating sounds implied by the image and ignoring
audio that conflicts with it~\cite{cmm,avhbench}. We ask whether this can be corrected cheaply, by
\emph{freezing the entire model} and training only a small per-modality variational information
bottleneck (VIB) inserted on the adapter token streams.

The training signal is a modality-conditional preference. From a clean clip we synthesize an
audio-swapped counterfactual (the original video with another clip's audio) and prefer the answer
consistent with the \emph{heard} event over the answer consistent with the \emph{seen} event. Because
both answers are scored on the same swapped clip, preferring the heard event requires using the audio.

We report three things. (i)~Applying DPO to this objective, while regularizing only the bottleneck's
information rate, collapses the model (Section~\ref{sec:collapse}); we attribute the collapse to
likelihood displacement and a missing output-distribution constraint. (ii)~Adding a chosen-likelihood
anchor and an explicit KL-to-base penalty prevents the catastrophic collapse, and an ablation shows
both terms are necessary (Section~\ref{sec:abl}). (iii)~With both terms, on Qwen3-Omni we measure a
modest, repeatable improvement on an audio--visual hallucination benchmark with capability preserved,
and we characterize the trained bottleneck (Section~\ref{sec:results}). We do not claim a
state-of-the-art result; the gains are small and the evaluation is on benchmark subsets.

\section{Related work}
\textbf{Preference optimization and its failure modes.} DPO~\cite{dpo} optimizes a log-sigmoid of the
difference between chosen and rejected log-probabilities. This difference can be increased by lowering
both, and the probability of the \emph{preferred} response can fall --- ``likelihood
displacement''~\cite{razin} --- worst when chosen and rejected are near-identical. DPO-Positive~\cite{dpop}
adds a penalty preventing the chosen log-probability from dropping below the reference. Unlike PPO, DPO's
KL constraint is implicit and acts only on the preference pairs~\cite{dpovsppo}.
\textbf{Multimodal preference optimization.} mDPO~\cite{mdpo} reports an analogous failure on multimodal
LLMs (the model ignores the image and the chosen likelihood drops) and adds a conditional term and a
reward anchor; our chosen anchor follows the same principle, with audio as the conditioning modality.
\textbf{Capability retention.} RLHF imposes an ``alignment tax''; explicit KL-to-reference penalties and
pretraining-mixing reduce it~\cite{instructgpt}, and proxy over-optimization past a KL budget degrades
true quality~\cite{gao}; selection should use diverse held-out evaluations~\cite{ivison}.
\textbf{Information bottleneck.} The variational IB~\cite{vib} trades sufficiency against a rate penalty;
Vittle~\cite{vittle} applies a VIB to multimodal-LLM hidden states. We use a per-modality VIB on adapter
outputs and report that the rate penalty, as commonly weighted, is weak (Section~\ref{sec:interp}).

\section{Method}
\subsection{Frozen AV-LLM with a per-modality VIB}
The base model is frozen. On each per-modality adapter output (the token sequence
$x\in\mathbb{R}^{T\times d}$ feeding the language model) we attach a VIB via a forward hook:
\begin{equation}
z=\mu(x)+\sigma(x)\odot\epsilon,\ \epsilon\sim\mathcal N(0,I);\qquad y=x+W_{\text{out}}z,
\end{equation}
with $W_{\text{out}}$ zero-initialized so $y=x$ at initialization. The per-token rate is
$\KL(q(z|x)\,\|\,\mathcal N(0,I))$. A bypass flag sets $y=x$, recovering the exact frozen base; we use it
for the DPO reference log-probabilities and the KL-to-base target without holding a second model. Only
the two VIBs (audio, vision) are trainable.

\subsection{Audio-swap preference and the naive objective}
\label{sec:collapse}
For a clip we build a multiple-choice question and an audio-swapped copy (video unchanged, audio replaced
by a different-category clip). The chosen letter is the heard event, the rejected letter the seen event,
both scored on the swapped clip. Let $c_p,r_p$ be the policy log-probs of the chosen/rejected letters
(bottleneck active) and $c_r,r_r$ the reference log-probs (bottleneck bypassed). The naive objective
regularizes only the latent rate $\KL_{\text{VIB}}$:
\begin{equation}
\mathcal L_0=-\log\sigma\!\big(\beta[(c_p-c_r)-(r_p-r_r)]\big)+\beta_{kl}\,\KL_{\text{VIB}}.
\label{eq:naive}
\end{equation}
Nothing constrains the output distribution against the base. Because $\mathcal L_0$ is a log-sigmoid of a
difference, it can be minimized by lowering both letter log-probs; with an always-on adapter and an
input-blind objective the probability mass moves to a degenerate constant answer.

\subsection{Anchored swap-DPO}
\label{sec:method-anchored}
We add two terms:
\begin{align}
\mathcal L=\ &-\log\sigma\!\big(\beta[(c_p-c_r)-(r_p-r_r)]\big)
 +\lambda_{a}\,\big(\!-\log\sigma(\beta(c_p-c_r)-\delta)\big)\nonumber\\
 &+\lambda_{kl}\,\KL\!\big(p_{\text{base}}(\cdot|x)\,\|\,p_\theta(\cdot|x)\big)+\beta_{kl}\,\KL_{\text{VIB}}.
\label{eq:anchored}
\end{align}
The \emph{chosen anchor} (second term, $\delta{=}0$)~\cite{mdpo,dpop} keeps the chosen log-prob at or above
the frozen base. The \emph{KL-to-base} term penalizes the policy for deviating from the base's answer
distribution on a set of general (non-swap) inputs $x$; because the base is the bypassed bottleneck,
$p_{\text{base}}$ is available without a second model. The general inputs span matched multiple-choice,
audio-presence, visual-presence, and open-ended prompts (the ``broad'' set).

\subsection{Monitoring and selection}
Per step we log the absolute chosen/rejected log-probs (not only their margin), $\KL_{\text{VIB}}$, the
KL-to-base, and the fraction of ``yes'' answers on a balanced yes/no probe (a constant-answer detector).
Checkpoints are saved periodically and selected on the held-out benchmarks, not the training proxy. The
selection rule maximizes AVHBench subject to capability guards (Section~\ref{sec:setup}).

\subsection{Implementation details (numerical)}
\label{sec:impl}
The Qwen-Omni backbones run in bfloat16. VideoLLaMA2 is distributed in float16 for inference but trained
in bfloat16; a differentiable float16 forward overflows attention, so we cast it to bfloat16. The VIB is
computed in float32 (bfloat16's mantissa is insufficient for the squared-mean term in the rate), its
encoder input is LayerNorm-normalized for backbones with large adapter activations, gradients are clipped
to norm $1.0$, and its linear layers are explicitly initialized (the VideoLLaMA2 loader disables the
default initialization process-wide). The Qwen backbones are unaffected by these guards.

\section{Experiments}
\subsection{Setup}
\label{sec:setup}
\textbf{Backbones:} Qwen3-Omni (primary), Qwen2.5-Omni, VideoLLaMA2.1-7B-AV\note{**}; all frozen, only the
two VIBs trained. \textbf{Training data:} audio-swapped AVE clips (300 pairs, 2 epochs, $\beta{=}0.1$,
$\lambda_a{=}1$, $\lambda_{kl}{=}2$, $\beta_{kl}{=}0.01$ unless stated). \textbf{Benchmarks (held-out,
$n{=}300$ subset each):} AVHBench~\cite{avhbench} (audio--visual hallucination, overall accuracy);
CMM~\cite{cmm} perception accuracy (PA; correctly affirms present content) and hallucination resistance
(HR; correctly denies absent content); DAVE~\cite{dave} audio--visual multiple choice. \textbf{Selection protocol (leakage control):} each benchmark's $300$ per-example records are split,
by a fixed seed, into a validation half and a disjoint test half; checkpoints (and the recipe and
$\beta_{kl}$) are selected on the validation half and \emph{all reported numbers are on the test half}.
The selection rule is the highest validation AVHBench subject to capability guards
$\text{PA}\ge\text{PA}_{\text{base}}-0.05$ and $\text{HR}\ge\text{HR}_{\text{base}}-0.05$ (relative to
each backbone's base, computed on validation). Reported metrics are therefore on $n{=}150$.

\subsection{The collapse and the anchor ablation (Qwen3-Omni)}
\label{sec:abl}
Table~\ref{tab:abl} reports the selected checkpoint for the four objectives; Table~\ref{tab:collapse}
reports the end-of-training (step 150) CMM axes to show the collapse dynamics. The naive objective
(Eq.~\ref{eq:naive}) reduces CMM perception (test) from $0.939$ to $0.015$ by step 150 and the model
answers a near-constant ``no'' (hallucination-resistance inflates to $1.000$ because denying everything
trivially avoids hallucination). The chosen anchor alone delays but does not prevent the perception loss
($0.227$ at step 150). The KL-to-base alone keeps perception near base but loses hallucination-resistance
($0.655$). Only with both terms does the selected checkpoint keep both axes within tolerance, and it
also attains the highest AVHBench. All four objectives drift if trained to step 150; selection is
mid-training.

\begin{table}[t]\centering\small
\caption{Anchor ablation on Qwen3-Omni (val-selected, test-reported, $n{=}150$). Selected checkpoint =
highest validation AVHBench with PA and HR within $0.05$ of base on validation. Higher is better.}
\label{tab:abl}
\begin{tabular}{lcccccc}
\toprule
Objective & $\lambda_a$ & $\lambda_{kl}$ & sel.\ step & AVHBench & CMM-PA & CMM-HR \\
\midrule
Frozen base & --- & --- & --- & 0.647 & 0.939 & 0.810 \\
Naive swap-DPO (Eq.~\ref{eq:naive}) & 0 & 0 & 30 & 0.673 & 0.939 & 0.774 \\
\quad + chosen anchor only & 1 & 0 & 30 & 0.660 & 0.939 & 0.798 \\
\quad + KL-to-base only & 0 & 2 & 60 & 0.647 & 0.924 & 0.810 \\
\quad + both (anchored, ``broad'') & 1 & 2 & 60 & \textbf{0.707} & 0.909 & 0.869 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[t]\centering\small
\caption{End-of-training (step 150) CMM axes on Qwen3-Omni (test split), showing the collapse without
both anchors. PA$\to$0.015 under naive DPO is a near-constant ``no'' (HR $\to$ 1.0 by denying everything).}
\label{tab:collapse}
\begin{tabular}{lcc}
\toprule
Objective (step 150) & CMM-PA & CMM-HR \\
\midrule
Naive swap-DPO & 0.015 & 1.000 \\
\quad + chosen anchor only & 0.227 & 0.893 \\
\quad + KL-to-base only & 0.879 & 0.655 \\
\quad + both (anchored) & 0.848 & 0.476 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Held recipe, seed variation, and rate sweep (Qwen3-Omni)}
\label{sec:results}
Table~\ref{tab:seeds} reports the anchored recipe across three seeds: the test AVHBench at the
validation-selected checkpoint is $0.694\pm0.023$ versus a test base of $0.647$, with every seed above base
and HR at or above base in all three. Table~\ref{tab:beta} sweeps the bottleneck rate weight $\beta_{kl}$:
the test AVHBench is comparable across $\{0.01,0.1,1.0\}$ (all $\approx0.70$), but the number of
checkpoints passing the validation guards increases with $\beta_{kl}$ (2/5, 4/5, 5/5), i.e. larger
$\beta_{kl}$ stabilizes training. This is consistent with the rate analysis in Section~\ref{sec:interp}
(the rate penalty at $0.01$ is weak). The apparent AVHBench advantage of $\beta_{kl}{=}0.1$ under full-set
(leaky) selection did not survive the val/test split.

\begin{table}[t]\centering\small
\caption{Anchored recipe across seeds (Qwen3-Omni, val-selected, test-reported). Base (test) AVHBench
$=0.647$, PA $=0.939$, HR $=0.810$.}
\label{tab:seeds}
\begin{tabular}{lcccc}
\toprule
seed & sel.\ step & AVHBench & CMM-PA & CMM-HR \\
\midrule
0 & 60 & 0.707 & 0.909 & 0.869 \\
1 & 90 & 0.667 & 0.864 & 0.810 \\
2 & 60 & 0.707 & 0.924 & 0.833 \\
\midrule
mean$\pm$std & --- & $0.694\pm0.023$ & $0.899\pm0.031$ & $0.837\pm0.030$ \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[t]\centering\small
\caption{Rate-weight sweep $\beta_{kl}$ (Qwen3-Omni, anchored ``broad'', seed 0; val-selected,
test-reported). ``guards pass'' counts how many of the 5 saved checkpoints satisfy both validation
guards. The test AVHBench is comparable across $\beta_{kl}$ (the apparent $\beta_{kl}{=}0.1$ advantage
under full-set selection was selection bias), but larger $\beta_{kl}$ is more stable.}
\label{tab:beta}
\begin{tabular}{lccccc}
\toprule
$\beta_{kl}$ & sel.\ step & AVHBench & CMM-PA & CMM-HR & guards pass \\
\midrule
0.01 & 60  & 0.707 & 0.909 & 0.869 & 2/5 \\
0.1  & 150 & 0.700 & 0.879 & 0.857 & 4/5 \\
1.0  & 150 & 0.707 & 0.909 & 0.869 & 5/5 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Multi-model results}
Table~\ref{tab:models} reports all three backbones. On Qwen2.5-Omni (test split) AVHBench rises from
$0.620$ to $0.673$ and perception is unchanged ($0.939$), while hallucination-resistance drifts down
slightly ($0.619$ to $0.583$); the relative guard (Section~\ref{sec:setup}) is required here because
Qwen2.5's base HR is below the absolute $0.70$ threshold used for Qwen3. VideoLLaMA2 entries are
base-only and marked \note{**}: a numerical bug (uninitialized bottleneck weights induced by the loader's
\texttt{disable\_torch\_init}, plus a float16 forward) prevented training; the fixes are in place but the
trained result is not yet validated.

\begin{table}[t]\centering\small
\caption{All backbones (AVHBench/CMM val-selected, test-reported; DAVE is the full-set base, not swept).
VideoLLaMA2 is base-only, pending validation (\note{**}).}
\label{tab:models}
\begin{tabular}{lcccc}
\toprule
& AVHBench & CMM-PA & CMM-HR & DAVE \\
\midrule
Qwen3-Omni base        & 0.647 & 0.939 & 0.810 & 0.380 \\
Qwen3-Omni anchored    & 0.707 & 0.909 & 0.869 & --- \\
Qwen2.5-Omni base      & 0.620 & 0.939 & 0.619 & 0.307 \\
Qwen2.5-Omni anchored  & 0.673 & 0.939 & 0.583 & --- \\
VideoLLaMA2 base\note{**}     & 0.677 & 0.720 & 0.087 & 0.357 \\
VideoLLaMA2 anchored\note{**} & ** & ** & ** & ** \\
\bottomrule
\end{tabular}
\end{table}

\subsection{What the trained bottleneck does}
\label{sec:interp}
A forward-pass probe of the held Qwen3-Omni checkpoint reads each VIB's per-token rate (KL in bits) and
edit magnitude ($\|W_{\text{out}}z\|/\|x\|$); Table~\ref{tab:interp} reports averages over 8 clips. The
vision VIB rewrites the vision tokens substantially ($\sim$59\% of each token's magnitude, spread across
all tokens) while the audio VIB is close to a pass-through (its small edit is a constant bias; its rate is
near zero). The grounding is thus achieved by reshaping the vision stream rather than by modifying audio;
because each VIB is per-modality, this transform is unconditional and the conflict resolution occurs in
the frozen language model. The vision rate ($\sim$640 bits/token) is large relative to the loss weight
$\beta_{kl}{=}0.01$ applied to the \emph{mean} rate, i.e. the bottleneck is weakly compressed at this
setting --- consistent with the $\beta_{kl}$ sweep, where larger weights stabilize training.

\begin{table}[t]\centering\small
\caption{Per-token VIB statistics of the held Qwen3-Omni checkpoint (8 clips).}
\label{tab:interp}
\begin{tabular}{lccc}
\toprule
modality & rate (bits/tok) & edit $\|\Delta\|/\|x\|$ & top-10\% of rate \\
\midrule
vision & $\sim$640 & $\sim$59\% & $\sim$27\% \\
audio  & $\sim$0.17 & $\sim$11\% (constant bias) & --- \\
\bottomrule
\end{tabular}
\end{table}

\section{Limitations}
The evaluation uses $300$-example subsets of each benchmark, split $150/150$ into a validation half (for
selection) and a test half (for reporting); test metrics are therefore on $n{=}150$ and are noisier than
the $\pm$ ranges alone suggest, and confidence intervals are over only three seeds for one recipe on one
backbone. The grounding gain is modest ($+0.047$ test AVHBench mean over seeds on Qwen3-Omni) and is the
validation-selected checkpoint, not a monotone training improvement --- all recipes drift if trained
longer. While the checkpoint, recipe and $\beta_{kl}$ are selected on the validation half, the
\emph{design} of the broad anchor was informed by observing the CMM failure on the held-out benchmarks ---
a residual form of test-informed design. On Qwen2.5-Omni hallucination-resistance drifts below base, so the
HR-preserving behavior on Qwen3-Omni does not transfer cleanly. VideoLLaMA2 results are not yet available
(\note{**}). DAVE was evaluated at the base only, not swept. The information-bottleneck framing is partly
nominal: at $\beta_{kl}{=}0.01$ the rate penalty is weak. We audit clip-id overlap between the AVE training
pool and the three benchmarks to check dataset-level leakage. Finally, the audio-swap signal is synthetic
and the benchmarks are yes/no-heavy, which interacts with the likelihood-displacement failure mode.

\section{Conclusion}
A frozen AV-LLM can be adapted toward audio grounding by training only a small per-modality information
bottleneck, but naive preference optimization collapses it. Anchoring the chosen likelihood to the base
and adding an explicit KL-to-base on general inputs prevents the collapse; an ablation shows both terms
are necessary. On Qwen3-Omni this yields a small, repeatable improvement on an audio--visual
hallucination benchmark with capability preserved, larger and more stable at higher bottleneck-rate
weights; results on Qwen2.5-Omni are partial and on VideoLLaMA2 pending. We release the diagnosis, the
recipe, the monitoring/selection protocol, and the numerical fixes.

{\small
\begin{thebibliography}{99}
\bibitem{dpo} Rafailov et al. Direct Preference Optimization. NeurIPS 2023. arXiv:2305.18290.
\bibitem{razin} Razin et al. Unintentional Unalignment: Likelihood Displacement in DPO. ICLR 2025. arXiv:2410.08847.
\bibitem{dpop} Pal et al. Smaug / DPO-Positive. 2024. arXiv:2402.13228.
\bibitem{mdpo} Wang et al. mDPO: Conditional Preference Optimization for Multimodal LLMs. EMNLP 2024. arXiv:2406.11839.
\bibitem{dpovsppo} Xu et al. Is DPO Superior to PPO for LLM Alignment? ICML 2024. arXiv:2404.10719.
\bibitem{instructgpt} Ouyang et al. Training Language Models to Follow Instructions with Human Feedback. NeurIPS 2022. arXiv:2203.02155.
\bibitem{gao} Gao, Schulman, Hilton. Scaling Laws for Reward Model Overoptimization. ICML 2023. arXiv:2210.10760.
\bibitem{ivison} Ivison et al. Unpacking DPO and PPO. NeurIPS 2024. arXiv:2406.09279.
\bibitem{vib} Alemi et al. Deep Variational Information Bottleneck. ICLR 2017. arXiv:1612.00410.
\bibitem{vittle} Oh et al. Visual Instruction Bottleneck Tuning (Vittle). NeurIPS 2025. arXiv:2505.13946.
\bibitem{avhbench} AVHBench: A Cross-Modal Hallucination Benchmark for Audio-Visual LLMs. 2024.
\bibitem{cmm} The Curse of Multi-Modalities: Hallucination in LMMs across Language, Visual, and Audio. 2024.
\bibitem{dave} DAVE: A Diagnostic Benchmark for Audio-Visual Evaluation. 2024.
\end{thebibliography}
}

\end{document}
